Bridging the Gap Between Visual and Analytical Machine Learning Testing

### 2023-09-25 report/esecfse24.tex
feat(report): 2023-09-25 23:22 updates

\subsection{Deeper Analysis of Notebooks}\label{sec:deep-dive}
% describe the steps used for the deeper analysis of NBs
% - we read the NBs in top-down fashion
% - reading the introductory md cell at the top often gives the context for the contents of the NB (eg. what type of ML model? which part of the ML lifecycle is this relevant for?)
% - reading the md cell above visualisation to understand context/motivation of visualisation
% - reading the source code that produces the visualisation
% - reading md cell after visualisation to understand the author's interpretation/observations from visualisation
% - reading the assertion
% - we fould 29 unique patterns from the 90 NBs; we highlight the 5 most interesting patterns in results

\section{Results}\label{sec:result}
This study identified 26 patterns of testing using visualisation and
assertion. This section highlights five interesting patterns in more
detail. We provide the complete catalogue of visualisation and
assertion pairs as supplemental material.

@@

% TODO try using captionof for the listing as well, so that the caption for the listing and figure align at the bottom
\begin{minipage}{0.5\textwidth}
  \begin{lstlisting}[language=Python]
assert accuracy_score(y_test, pred) > 0.95
  \end{lstlisting}
  \captionof{lstlisting}{Assertion on the accuracy of the ML model.}
  \label{lst:svm}
\end{minipage}
\begin{minipage}{0.5\textwidth}
  \includegraphics[width=\linewidth]{../catalogue/select-04.png}
  \captionof{figure}{Visualisation of the decision boundary of a SVM.}
  \label{fig:svm}
\end{minipage}

% is the data from the training or testing set?
% what is the info obtained from the visualisation?
% what is the assertion doing? what is the info obtained here?
% does the assertion reflect the info obtained from the visualisation?

Figure~\ref{fig:svm} presents a visualisation obtained from a notebook that performs binary classification using Support Vector Machines (SVM). The model is trained on the \texttt{make\_circles}~\footnote{REFME} dataset obtained from scikit-learn~\footnote{REFME}. The visualisation shows the two classes present in the training set, and the decision boundary of the model.

Listing~\ref{lst:svm} shows the corresponding assertion that was defined below Figure~\ref{fig:svm} in the notebook. The assertion checks if the model accuracy is above a threshold specified by the author of the notebook.

@@

Figures~\ref{fig:football} and~\ref{fig:tennisball} are obtained from a notebook that uses a pre-trained image classifier from GluonCV library~\footnote{https://cv.gluon.ai}, to identify images that contain tennis balls.

Listing~\ref{lst:football} presents the assertion related to Figure~\ref{fig:football}. The assertion checks if the probability estimate of the model for the given input image, is close to the specified value with a precision of up to three decimal places. Since the model is trained to detect tennis ball, and the input image is of a football, the specified probability estimate is very low.

@@

In contrast, Listing~\ref{lst:tennisball} specifies a much higher probability estimate since the input image contains tennis balls.

@@

Figure~\ref{fig:r2} shows a visualisation obtained from a notebook that trains a RNN to predict the levels of Nitrogen Dioxide ($NO_2$) in the next 24 hour window. The model is trained using time-series data consisting of historical readings of $NO_2$ and meteorological data such as temparate, humidity, wind speed, and such.

% I want to say that the spread of the points should be close to the diagonal to indicate a good fit which is not the case here.
The visualisation shows a scatterplot of the real and predicted values of $NO_2$ from the test set. The scatterplot shows a linear relationship between the ground truth and predictions indicating that the model learned something during training. However, the spread of the points is wide indicating that it can be improved with more fine tuning.

Listing~\ref{lst:r2} presents the corresponding analytical test which was defined below Figure~\ref{fig:r2}. The Coefficient of Determination ($R^2$) is calculated using the \texttt{r2\_score} method from scikit-learn~\footnote{REFME}. The assertion checks that the $R^2$ is higher than a specified threshold.

@@

\subsection{Time-series Prediction}

% TODO this is condition0! find an alternative!

\begin{minipage}{0.5\textwidth}
  \begin{lstlisting}[language=Python]
# Result of the wfc function for the best choice of s
assert np.isclose(best_wfv, 7685.546728971963, 0.001)
  \end{lstlisting}
\end{minipage}
\begin{minipage}{0.5\textwidth}
  \includegraphics[width=\linewidth]{../catalogue/select-117b.png}
\end{minipage}

>>>>

\subsection{Deeper Analysis of Notebooks}\label{sec:deep-dive}

We qualitatively assess the comments made for all notebooks during the analysis conducted in phase two of the study, as described in Section~\ref{sec:identify-related-pairs}. We perform an in-depth analysis of 90 short-listed notebooks to identify unique ML testing tasks. The in-depth analysis was conducted by reading and understanding the entire contents of the notebooks in a top-down order. During the analysis, we note several data points such as the primary objective of the notebook, the type of ML problem, the type of ML model and the type of data. We manually read and comprehend each code cell in the notebooks. When present, we also read the markdown cells above and below code cells to gain further context from the notebook author. For instance, we find that markdown cells above a visualisation cell typically motivates the need for the visualisation. A markdown cell directly below a visualisation often records the observations made by the notebook author from the visualisation.

\section{Results}\label{sec:result}
This study identified 26 patterns of testing using visualisation and assertion. This section highlights five interesting patterns in more detail. We plan to release the entire catalogue of patterns publicly in the near future.

@@

\begin{minipage}{0.5\textwidth}
  \begin{lstlisting}[language=Python]
assert accuracy_score(y_test, pred) > 0.95
  \end{lstlisting}
  \captionof{lstlisting}{Assertion on the accuracy of the ML model.}
  \label{lst:svm}
\end{minipage}
\begin{minipage}{0.5\textwidth}
  \includegraphics[width=\linewidth]{../catalogue/select-04-linear.png}
  \captionof{figure}{Visualisation of the decision boundary of a SVM with a linear kernel.}

  \includegraphics[width=\linewidth]{../catalogue/select-04-rbf.png}
  \captionof{figure}{Visualisation of the decision boundary of a SVM with a RBF kernel.}
  \label{fig:svm}
\end{minipage}

Figures~\ref{fig:svm} presents visualisations obtained from a notebook that performs binary classification using Support Vector Machines (SVM). The model is trained on the \texttt{make\_circles}~\footnote{REFME} dataset obtained from scikit-learn~\footnote{REFME}. The visualisations show the two classes present in the training set, and the decision boundary of the model when trained first using a linear kernel and later with a RBF kernel. Listing~\ref{lst:svm} shows the corresponding assertion that was defined below the visualisations presented in Figure~\ref{fig:svm}. The assertion checks if the model accuracy is above a threshold specified by the author of the notebook.

@@

@@

Figures~\ref{fig:football} and~\ref{fig:tennisball} are obtained from a notebook that uses a pre-trained image classifier from GluonCV library~\footnote{https://cv.gluon.ai}, to identify images that contain tennis balls.The visualisations are of two random input images, one from each class in the dataset. Listing~\ref{lst:football} presents the assertion related to Figure~\ref{fig:football}. The assertion checks if the probability estimate of the model for the given input image, is close to the specified value with a precision of up to three decimal places. Since the model is trained to detect tennis ball, and the input image contains a footbal, the specified probability estimate is very low. In contrast, Listing~\ref{lst:tennisball} specifies a much higher probability estimate since the input image contains tennis balls.

% NOTE when working with images, we tend to see that the input image is visualised (so human can manually validate), something to do with expensive labelling?
% NOTE is this a good pattern? this test does not generalise across all possible input images?

@@

Figure~\ref{fig:r2} shows a visualisation obtained from a notebook that trains a RNN to predict the levels of Nitrogen Dioxide ($NO_2$) in the next 24 hour window. The model is trained using time-series data consisting of historical readings of $NO_2$ and meteorological data such as temparate, humidity, wind speed, and such. The visualisation shows a scatterplot of the real and predicted values of $NO_2$ from the test set. The scatterplot shows a linear relationship between the ground truth and predictions indicating that the model learned something during training. However, the spread of the points is wide indicating that it can be improved with more fine tuning. Listing~\ref{lst:r2} presents the corresponding analytical test which was defined below Figure~\ref{fig:r2}. The Coefficient of Determination ($R^2$) is calculated using the \texttt{r2\_score} method from scikit-learn. The assertion checks that the $R^2$ is higher than a specified threshold.

@@

\subsection{Normalization}

\begin{minipage}{0.5\textwidth}
  \begin{lstlisting}[language=Python]
assert normalized_data[normalized_data.argmax()] < 10
assert normalized_data[normalized_data.argmin()] > -7
  \end{lstlisting}
  \captionof{lstlisting}{Assertion to check that the min and max of a feature fall within specified thresholds derived from the visualisation presented in Figure~\ref{fig:normal}.}
  \label{lst:normal}
\end{minipage}
\begin{minipage}{0.5\textwidth}
  \includegraphics[width=\linewidth]{../catalogue/select-65b.png}
  \captionof{figure}{Histogram showing the distribution of a feature before and after normalisation.}
  \label{fig:normal}
\end{minipage}
